---
title: "What is an R Markdown"
author: "Jill Marzolino"
date: "October 4, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Problem 2: Calculating Fst

Load the file combined_YRICEU.txt as shown below, which contains both the CEU and YRI genotype counts
for each SNP.

```{r}
library(eve102)
d_comb <- read.table(eve102_data("combined_YRICEU.txt"), header=TRUE)
head(d_comb)
```

###Question 1:
With this data, calculate the mean expected heterozygosity (i.e. from allele frequencies) in
European and African samples (CEU and YRI, respectively) separately.

CEU
```{r}
freqs_CEUA <- (d_comb$AA_CEU+d_comb$Aa_CEU/2)/d_comb$total_CEU
freqs_CEUa <- (d_comb$aa_CEU+d_comb$Aa_CEU/2)/d_comb$total_CEU
Hs_CEU <- mean(2*(freqs_CEUA)*(freqs_CEUa))
```

YRI
```{r}
freqs_YRIA <- (d_comb$AA_YRI+d_comb$Aa_YRI/2)/d_comb$total_YRI
freqs_YRIa <- (d_comb$aa_YRI+d_comb$Aa_YRI/2)/d_comb$total_YRI
Hs_YRI <- mean(2*(freqs_YRIA)*(freqs_YRIa))
```

###Question 2:
Calculate Fst for the European population relative to the combined frequency. 
average heterozygosity across loci and calculate Fst

Total Heterozygosity

find allele freq of ceu (p1 and q1) find allele freq of combined (pt and qt), then divide ps and qs by 2 and then do 2pq
average ps and qs
```{r}
p_tot <- (freqs_YRIA)+(freqs_CEUA)/2
q_tot <- (freqs_YRIa)+(freqs_CEUa)/2
H_tot <- mean(2*q_tot*p_tot)
```


Fst for CEU
```{r}
Fst_CEU <-1-(Hs_CEU/H_tot)
Fst_CEU
```


###Question 3:
Calculate Fst for the African population relative to the combined frequency.

Fst for YRI
```{r}
Fst_YRI <- 1-(Hs_YRI/H_tot)
Fst_CEU
```


###Question 4: 
Take the average of these two Fsts.

```{r}
Fst_comb <- (Fst_CEU+Fst_YRI)/2
Fst_comb
```

###Question 5: 
Your calculated Fst is the inbreeding coefficient, giving the average reduction of heterozygosity in the sub-populations relative to the total population. 
The expected proportion of heterozygotes with inbreeding is given by the generalized Hardy-Weinberg equation (equation 3, page 9 in the course notes: [(1-F)2pq].
Using the generalized Hardy-Weinberg equilibrium, and your calculated Fst,
plot a line for the expected number of heterozygotes as a function of the allele frequency 
x=combined freq of AA, y=expected het (gen_HW)
```{r}
Fst_comb
gen_HW <- (1-Fst_comb)*2*p_tot*q_tot
plot(p_tot,gen_HW,pch=18,cex=0.5,col="green",xlim=c(0,1.55),ylim=c(0,0.6),xlab="allele frequency",ylab="expected heterozygotes")
```