---
title: "EVE102, Homework 2"
---

```{R, include=FALSE}
knitr::opts_chunk$set(prompt=FALSE, comment="", fig.width=6, fig.height=6)
```

## Section 1: Wright-Fisher Population Simulation

In this exercise, you'll use some code that simulates a Wright-Fisher
population to look how heterozygosity decays through random genetic drift.
We've code the Wright-Fisher simulation as simply as possible so that if you're
curious you can learn how it works and if you like modify it.

You'll need to run the following code block to load the `wf()` function into R.

```{R}
wf <- function(N, ngens, p0=1/3, mu=0) {
 N <- 2*N  # diploid adjustment
 # initialize an empty matrix
 gns <- matrix(NA, nrow=ngens, ncol=N)
 # initialize the first generation, with two alleles, one at freq
 # p0
 alleles <- 2
 gns[1, ] <- sample(1:2, N, replace=TRUE, prob=c(p0, 1-p0))
 for (i in 2:ngens) {
  gns[i, ] <- gns[i-1, sample(1:N, N, replace=TRUE)]
  if (mu > 0) {
    # add mutations to this generation
    muts <- rbinom(N, 1, prob=mu)
    new_alleles <- sum(muts)
    alleles <- alleles + new_alleles
    if (new_alleles) {
      # there are mutations, add to population.
      gns[i, ] <- ifelse(muts, sample(alleles), gns[i, ])
    }
  }
 }
 gns
}
```

## The decay of heterozygosity

**Question 1**:

Fill in the following function `het()`  to create a function that takes a set
of alleles like 1, 1, 1, 1, 2, 2, 3 and calculates the heterozygosity. To do
this, you'll need to use the functions `table()`, which takes a vector and
returns the counts. For example:

```{R}
alleles <- c(1, 1, 1, 1, 2, 2, 3)
table(alleles)
```

We use this function because we could have many alleles in the model, and
cannot assume biallelic loci. Then, use `sum()` to get the total number of
alleles to calculate the allele frequencies of each allele. Then use what you
know about homozygosity and heterozygosity (this is very similar to Question 1)
in the notes) to calculate the heterozygosity. Do this in the following
function, where `x` is the vector of alleles (like `alleles` above):

```{R}
het <- function(x) {
  allelefreq <- table(x)
  1-sum((allelefreq/sum(allelefreq))^2)
  #heterozygosity is one minus the sum of all the homozygous probabilities
  #each homozygosity is the allele frequency (number of that allele out of the total number) squared
}
```


```{R,echo=FALSE}
# not shown 
het <- function(x) {
 tbl <- table(x)
 1 - sum((tbl/sum(tbl))^2)
}
```

You can test your function with the vector `alleles` -- you should get:

```{r}
het(alleles)
```

**Question 2**: In this question, we'll use the `wf()` simulation function
*with no mutation* to see the decay of heterozygosity. Since drift is a
stochastic process, we need to average over many replicate populations to see
the average decay. First, to familiarize yourself with the `wf()` function, run
it with $N=100$ for 250 generations (using `ngens=250`).

`wf()` returns a matrix, where each row is a generation and contains the
alleles of each of the individuals. For example:

```{r}
set.seed(0)  # ensures your random results are same as discussed below
wf(N=10, ngens=2)
```
Here, in the first generation, individual 1 has allele 1, individual 2 has
allele 2, individual 3 has allele 2, etc.

We need to summarize the heterozygosity *per generation*. This involves passing
each row to the `het()` function you wrote above, like:

```{R}
sim <- wf(N=10, ngens=2)
het(sim[1, ])
het(sim[2, ])
```
etc. However, if we simulate 250 generations, we don't want to have to do this 250 times. So, we use a function called `apply()`, which takes three arguments: a matrix, either 1 or 2 (1 means summarize over rows, 2 means summarize over columns), and a function that is used to summarize either rows or columns.

For example, imagine we have the matrix `d` below. We can use `apply(d, 1,
sum)` to calculate the sum of each row.

```{R}
d <- matrix(c(3, 1, 1, 2,
              1, 1, 3, 1), byrow=TRUE, nrow=2)
apply(d, 1, sum)
```

Using this approach, we can calculate heterozygosity across generations.
`apply()` will return a vector: **use this vector to plot the decay of
heterozygosity across the simulations you ran above**.

These are repeats of what is increasingly simplified code chunks, but I decided to keep them because they effectively illustrate the randomness of drift with three iterations. 
```{R}
A <- wf(N=100,ngens = 250)
B <- apply(A,1,het)
plot(1:250,B, xlab = "generations" , ylab = "heterozygosity",pch=18, col="navyblue")
```

```{r}
B <- apply((wf(N=100,ngens = 250)),1,het)
plot(1:250,B, xlab = "generations" , ylab = "heterozygosity",pch=18, col="navyblue")
```

```{r}
plot(1:250,(apply((wf(N=100,ngens = 250)),1,het)), xlab = "generations" , ylab = "heterozygosity",pch=18, col="navyblue")
```

**Question 3**:

Drift is a stochastic process; each of these simulations is variable. We want
to replicate the population many times and take the average of all of these to
understand generally how heterozygosity declines.

Since we summarize each replicate by calculating heterozygosity, we don't need
to store each full population of $N$ individuals across the `ngnens`. So, we
need to write a function that *wraps* `wf()` but returns a vector of
heterozygosity across the generations, as you did above using `apply()`. Fill
in the function below to do this (we've set up the part that wraps `wf()` for
you:

```{R}
het_sim <- function(N, ngens, p0=1/3, mu=0) {
  sim <- wf(N=N, ngens=ngens, p0=p0, mu=mu)
  # ADD heterozygosity calculation here, and return vector
  # of heterozygosity per generation (uses apply())
  (sum(apply(sim,1,het)))/ngens
  #takes each row of "sim" and runs het function on it, add all of those together and divide by the total number of generations
  #yields a vector of heterozygosity per generation
}
```

```{r}
het_sim <- function(N, ngens, p0=1/3, mu=0) {
  sim <- wf(N=N, ngens=ngens, p0=p0, mu=mu)
  # ADD heterozygosity calculation here, and return vector
  # of heterozygosity per generation (uses apply())
  apply(sim,1,het)
  #takes each row of "sim" and runs het function on it
  #yields a vector of heterozygosity for every generation
}
```


Using this `het_sim()` function, we're now going to replicate many populations
and summarize their heterozygosity. R has a useful function for this:
`replicate()`, which takes in the number of replicates and the function to run,
and returns the results in a nice simplified matrix. Here's an example of how
to replicate this process 2 times:

```{R}
reps <- replicate(2, het_sim(100, 200))
```

**Replicate the simulation 30 times, for $N=100$, for 500 generations**

```{R}
reps30 <- replicate(30,het_sim(100,500))
```

**Now, use `rowMeans()` to take the average heterozygosity per generation,
across all simulations.** `rowMeans()` takes in the matrix created above, and
averages across rows (replicate simulations).

```{R}
Y <- rowMeans(reps30)
```

**Finally, plot these average heterozygosity values.**

```{R}
plot(1:500,Y, col="purple",pch=18,xlab="generations",ylab="average heterozygosity")
```

**Question 4**: Using your population size ($N=100$) calculate the decay of
heterozygosity through time, from its original value ($p = 1/3$), using the
exponential approximation. Using `seq()` to generate a vector sequence of
generations, and `lines()` to overlay this approximation, overlay a line of
expected theoretical decay on your last plot. 

```{R}
#H_t = H_o*exp(-t/(2*N))
t <- 500
N <- 100
p <- 1/3
H_o <- 2*p*(1-p)
H_t <- H_o*exp(-(seq(1,500))/(2*N))
plot(1:500,Y, col="purple",pch=18,xlab="generations",ylab="average heterozygosity")
lines(seq(1,500),H_t,col="red")
```

## Mutation-Drift Balance

Now, we're going to add in mutation (albeit at an artificially high rate to
make the simulations easier).

**Question 1**: Using the `mu` argument of your `het_sim()` function, add in
mutations at a rate of $\mu = 1e-3$. Increase the number of generations to 400.
Replicate the population as above (but with $N=100$ and `ngens=400`), and
summarize mean heterozygosity. Create a plot as before.

```{R}
# your simulation and plot code here
repsmu <- replicate(30,het_sim(N=100,ngens=400,mu=exp(-3)))
plot(1:400,rowMeans(repsmu), col="forestgreen",pch=18,xlab="generations",ylab="average heterozygosity")
```

**Question 2**: Using the function `abline()`, which takes an argument `h` for
a horizontal line, add in a vertical line showing what you'd expect for under
mutation-selection balance. 

```{R}
#equilibrium heterozygosity in a population at equilibrium between mutation and drift
#H=(4*N*mu)/(1+(4*N*mu))
N <- 100
mu <- exp(-3)
H <- (4*N*mu)/(1+(4*N*mu))
plot(1:400,rowMeans(repsmu), col="forestgreen",pch=18,xlab="generations",ylab="average heterozygosity",ylim = c(.45,1))
abline(h=H)
```

**Question 3**: Redo the above calculation where $\mu=1e-2$, and create a plot
and expected line under mutation-drift balance.

```{r}
repsmu2 <- replicate(30,het_sim(N=100,ngens=400,mu=exp(-2)))
plot(1:400,rowMeans(repsmu2),col="red",pch=18,xlab="generations",ylab="average heterozygosity",ylim = c(.45,1))
mu2 <- exp(-2)
H2 <- (4*N*mu2)/(1+(4*N*mu2))
abline(h=H2)
```